Leaf traits drive differences in biomass partitioning among major plant functional types

### 2015-01-12 manuscript.Rnw
Text editing; more refs.

<<preload, echo=FALSE>>=
library(knitr)
opts_chunk$set(echo=FALSE)
source("R/functions-ms.R")
baad <- readRDS("downloads/baad.rds")
dataauthors <- author_details(baad, "data/contact.csv")
@

@@

\textbf{The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time of carbon in ecosystems \citep{friend2014}, predictions of global carbon storage \citep{ise2010}, and the CO\textsubscript{2} response of net primary production \citep{dekauwe2014}. Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking. Using a new global database on biomass partitioning from over 175 studies, we here show a strong effect of PFT on partitioning between stems and leaves. We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass. Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum \citep{wrightetal2004}. In addition, we show how a second trait, leaf area per stem basal area, contributes additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs. Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
}

@@

The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation\cite{ise2010}. Globally, forests store approximately 360Pg of carbon\citep{panetal2011}, equivalent to about 40years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}. Biomass stored in woody stems is unproductive but has a long residence time \citep{luyssaert2008}, whereas leaf biomass is enhances photosynthesis but turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \cite{ryanlaw2005}. Reducing uncertainties about biomass partitioning and carbon residence times in GVMs is therefore a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014}.

Perhaps and the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising world vegetation. While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots) and the quality with which these are constructed. The approach taken by most GVMs for handling this functional diversity, is to consider only a few archetypal PFTs \citep{harrisonetal2010}, assumed to differ in key physiological attributes. While it is common to assume differences in biomass partitioning between PFTs, these estimates are based on limited empirical data. Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm}) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared.

Here, we use a unique, new database\citep{Falster2015} (Fig. 1) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms. The dataset includes over 21,000 observations of biomass and size metrics for individual plants, contributed by 175 studies across nine vegetation types (Fig. S1). Previous work based on whole stands\citep{oneilldeangelis1981,enquistniklas2002,reich2014} and very small plants\citep{poorter2012}, reported differences among angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF). Here we extend this finding to show that LMF differs not only across the angiosperm-gymnosperm divide, but also within angiosperms, at the level of individual plants, and across the entire size range of woody plants in natural forests (0.1-100m) (Fig. 2a).

Moreover, we show that differences in LMF among woody PFTs arise due to differences in leaf morphology, not in the amount of leaf area held on the plant (Fig. 2b-c). Previously, it was unclear whether PFTs with higher leaf biomass also had higher total leaf area. Leaf area is relevant because it drives total light interception and thus photosynthesis. Several classic studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}, suggesting a growth advantage. We found no evidence for this, instead we found that the three PFTS maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR). Because of allometric relationships between the studied variables \citep{niklas1994}, there was a strong effect of plant height on LMF (Fig. 2a and S2). We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}. We then found that LMF at a common height was proportional to the leaf construction cost, as measured by leaf mass per area (LMA) (Fig. 2b). The consequence of this is that the leaf area ratio (the amount of leaf area supported per unit above-ground biomass), was invariant between PFTs, because $\textrm{LAR} = LMF / \textrm{LMA}$ (Fig. 2c and S3).

Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem; quantifying the amount of light intercepting tissue relative to the amount of tissue available to transport water through the stem. This measure has long been assumed to ..(reason why it's important)... Mirroring the results for LMF, we find that the amount of leaf mass per unit stem area differs among the three PFTs, that these differences also relate to LMA (Fig. 2d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area (Fig. 2e).

The above results highlight how differences in LMA -- a central trait of the leaf economics spectrum\citep{wrightetal2004} -- drives differences in biomass partitioning among dominant woody PFTs. Based on a XXX statistical model, we found that LMA alone could explain almost as much  (?more?) variation as PFT, even though LMA only captures aspects of leaf physiology. These results suggest promising avenues for parameterizing and simplifying GVMs.

Given these results, we were also interested to know whether additional variation in biomass partitioning could be explained by either climate, within PFT variation in LMA or by other traits. Despite the size of our dataset, the strong size-dependence of LMF greatly hinders comparisons among species and across climatic gradients, due to the small sample size within each species or site. Thus, to answer these additional questions we found it helpful to decompose LMF as the product, 
\begin{equation}\label{eq:M1}\textrm{LMF} = \left(\frac{M_F}{A_S}\right) \left(\frac{A_S}{M_T}\right).\end{equation}
An advantage of eq. \ref{eq:M1} is that the second term absorbs nearly all of the size related variation in LMF (Tables 1-2) and is otherwise fairly conserved between species\citep{chave2005} and across PFTs (Fig. S4). As shown in Fig. 2b and Fig. 2d, the remaining term $M_F/A_S$ exhibits comparable differences to LMF, but unlike $M_F/A_S$ is independent of plant height (Tables 1-2). $M_F/A_S$ is thus a useful of LMF that can be compared across species and climates.

@@

The finding that climate does not appreciably affect biomass partitioning is good news for models that parameterize the globe by PFT. Yet, these results contrast those of a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}. One explanation for these differences is that our measurements were taken on individual plants, whereas in the other study they were on whole stands. This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants. With the caveats that there is still variation between species and studies unaccounted for (Fig. S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT seem stronger than any direct climate effect on biomass partitioning within individual plants (Tables 1 and 2).

@@

Overall, our results establish general patterns about plant construction and thus lay an empirical base against which models can be benchmarked. We found that PFTs differed in LMF, and this was proportional to LMA. It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning. A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions. A first approximation within GVMs would be to keep leaf area to stem cross-sectional area constant for a given PFT, and vary biomass allocation accordingly. LMA, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem biomass. An example of such a model is LPJ \citep{sitchetal2003}, where the ratio of leaf area to sapwood area is a constant parameter, but LMA differs between PFTs. More fundamentally, our results suggest PFTs defined on the basis of two key traits, LMA and $A_F/A_S$ would better capture patterns of biomass partitioning than existing groups, which only account for differences in one of the two traits identified here. In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{Falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz_towards_2012, dekauwe2014} of GVMs against empirical data.

@@

We further decomposed LMF and LAR by considering the ratio of leaf mass (or area) to plant basal area, and the ratio of above-ground biomass to plant basal area. For LMF,

$$\textrm{LMF} = \frac{M_F}{M_T} = \left(\frac{M_F}{A_S}\right) \left(\frac{M_T}{A_S}\right)^{-1}.$$

And similarly for LAR, by replacing $M_F$ with $A_F$. We were unable to further decompose $A_S/M_T$ into wood volume and wood density components (cf. \citep{chave2005}) because wood density data were available only for a small subset of data.

@@

Variance explained by quantitative climate variables (MAP and MAT) were tested with GAMs where variables were sequentially added to the model, and the explained variance ($R^2$) calculated. For the test of biome effects on biomass partitioning, we used a linear mixed-effects model (because two factors and their interactions were tested), again with species-dataset as the random effect. We calculated the $R^2$ for linear mixed-effects models as reported by \citep{nakagawa2013} for the fixed effects only.

@@

Figure 1a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
    \caption{\textbf{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning}. a) LMF for each PFT within three different biomes.  No leaf area estimates were available for boreal deciduous angiosperms. Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect. Letters denote significant differences (at $\alpha = 0.05$). b-c) Relationships between climate and LMF within each PFT. Solid lines are generalized additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
    \label{fig:figure3}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure4.pdf}
     \caption{\textbf{Leaf and stem traits capture variation in biomass partitioning across species.} Individual plant data were averaged by species and study combinations. Lines are linear regressions with 95\% confidence intervals. Both regressions were significant ($P \textless 0.01$). $R^2$ for fitted relationships are 0.28 and XXX in the right panel.}
  \label{fig:figure4}
\end{figure}

@@

\clearpage

@@

\subsection*{Methods S1 - Biomass partitioning terminology}
\label{sec:partterm}

The terms 'partitioning' and 'allocation' have been used in various ways, confusing comparisons between studies \citep{litton2007}. Here, we define biomass partitioning as the actual distribution of biomass between compartments (e.g., leaves vs. stems), and biomass allocation as the proportion of net primary production (NPP) that is allocated to some compartment. The two processes are different because of continuous turnover of biomass, which differs strongly between compartments. We may write (cf. \citep{McMurtrieWolf1983}),

@@

\subsection*{Methods S2 - Predicting basal diameter}
\label{sec:predas}

>>>>

<<preload, echo=FALSE>>=
library(knitr)
opts_chunk$set(echo=FALSE)
source("R/functions-ms.R")
baad <- readRDS("downloads/baad.rds")
@

@@

\textbf{The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time of carbon in ecosystems \citep{friend2014}, predictions of global carbon storage \citep{ise2010}, and the CO\textsubscript{2} response of net primary production \citep{dekauwe2014}. Although GVMs typically predict that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking. Using a new global database on biomass partitioning from over 175 studies, we here show a strong effect of PFT on partitioning between stems and leaves. We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass. Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum \citep{wrightetal2004}. In addition, we show how a second trait, leaf area per stem basal area, explains additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs. Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
}

@@

The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation\cite{ise2010}. Globally, forests store approximately 360Pg of carbon\citep{panetal2011} in living biomass, equivalent to about 40years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}. Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass determines turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \cite{ryanlaw2005}. Reducing uncertainties about biomass partitioning and carbon residence times in GVMs is therefore a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014}.

Perhaps the biggest challenge for GVMs is to capture the combined responses of the more than 250,000 plant species comprising the world's vegetation. While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of energy invested in different tissues (leaves, stems, roots). The approach taken by most GVMs for dealing with this functional diversity is to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010}, assumed to differ in key physiological attributes. While GMVs predict differences in biomass partitioning between PFTs (Methods \ref{sec:gvmsummary}), these predictions are constrained by limited empirical data. Moreover, there is little consensus on how biomass partitioning and allocation (see Methods \ref{sec:partterm} for terminology) should be modelled in GVMs \citep{franklin2012,dekauwe2014,friend2014}. These shortcomings largely reflect the lack of suitable datasets of global scope with which models can be tested, constrained and compared.

Here, we use a unique, new database\citep{Falster2015} (Fig. 1) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms. The dataset includes over 21,000 observations of biomass and size metrics for individual plants, contributed by 175 studies across nine vegetation types (Fig. S1). Previous work based on whole stands\citep{oneilldeangelis1981,enquistniklas2002,reich2014} and very small plants\citep{poorter2012}, reported differences between angiosperms and gymnosperms in the amount of leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF). Here we extend this finding to show that LMF differs not only across the angiosperm-gymnosperm divide, but also within angiosperms, at the level of individual plants, and across the entire size range of woody plants in natural forests (0.1-100m) (Fig. 2a).

Moreover, we show that differences in LMF among woody PFTs arise due to differences in leaf morphology, but they converge in the amount of leaf area held on the plant (Fig. 2b-c). Previously, it was unclear whether PFTs with higher leaf biomass also had higher total leaf area. Leaf area is relevant because it drives total light interception and thus photosynthesis. Several classic studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}. We found no evidence for this, instead we found that the three PFTS maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR). Because of allometric relationships between the studied variables \citep{niklas1994}, there was a strong effect of plant height on LMF (Fig. 2a and S2). We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}. We then found that LMF at a common height was proportional to the leaf construction cost, as measured by leaf mass per area (LMA) (Fig. 2b). As a consequence, LAR was invariant between PFTs, because $\textrm{LAR} = LMF / \textrm{LMA}$ (Fig. 2c and S3).

Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem ($A_S$), a quantity closely related to hydraulic supply capacity \cite{enquist2003}. This ratio is also important because plant leaf mass is frequently estimated from $A_S$ with allometric equations \cite{shinozaki1964b, chave2005}.  Mirroring the results for LMF, we find that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig. 2d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig. 2e).

The above results highlight how differences in LMA -- a central trait of the leaf economics spectrum\citep{wrightetal2004} -- drives differences in biomass partitioning among dominant woody PFTs. Based on a XXX statistical model, we found that LMA alone could explain almost as much  (?more?) variation as PFT, even though LMA only captures aspects of leaf morphology. These results suggest promising avenues for parameterizing and simplifying GVMs.

Given these results, we were also interested to know whether additional variation in biomass partitioning could be explained by either climate, within PFT variation in LMA or by other traits. Despite the size of our dataset, the strong size-dependence of LMF greatly hinders comparisons among species and across climatic gradients, due to the small sample size within each species or site. Thus, to answer these additional questions we found it helpful to decompose LMF as the product,

\begin{equation}\label{eq:M1}
\textrm{LMF} = \left(\frac{M_F}{A_S}\right) \left(\frac{A_S}{M_T}\right).
\end{equation}

An advantage of eq. \ref{eq:M1} is that the second term absorbs nearly all of the size related variation in LMF (Tables 1-2) and is otherwise fairly conserved between species\citep{chave2005} and across PFTs (Fig. S4). As shown in Fig. 2b and Fig. 2d, the remaining term $M_F/A_S$ exhibits comparable differences to LMF, but unlike $M_F/A_S$ is independent of plant height (Tables 1-2). $M_F/A_S$ is thus a useful proxy for LMF that can be compared across species and climates.

@@

The finding that climate does not appreciably affect biomass partitioning is good news for models that parameterize the globe by PFT. Yet, these results contrast those of a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}. One explanation for these differences is that our measurements were taken on individual plants, whereas in the other study they were on whole stands. This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants. With the caveats that there is still variation between species and studies unaccounted for (Fig. S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables 1 and 2).

@@

Overall, our results establish general patterns about plant construction and thus lay an empirical base against which models can be benchmarked. We found that PFTs differed in LMF, and this was proportional to LMA. It thus seems possible to integrate a key leaf trait with whole-plant modelling of biomass partitioning. A recent study compared allocation routines in a number of leading ecosystem models \citep{dekauwe2014}, and recommended constraining allocation by observed biomass fractions instead of using constant allocation fractions. A first approximation within GVMs would be to keep leaf area to stem cross-sectional area constant for a given PFT, and vary biomass allocation accordingly. LMA, already a parameter in most GVMs, then gives the ratio of leaf biomass to stem biomass. An example of such a model is LPJ \citep{sitchetal2003}, where the ratio of leaf area to sapwood area is a constant parameter, but LMA differs between PFTs.

More fundamentally, our results suggest PFTs defined on the basis of two key traits, LMA and $A_F/A_S$ would better capture patterns of biomass partitioning than existing groups, which only account for differences in one of the two traits identified here. In any case, the growing availability of large datasets on stand biomass and individual plant construction  \citep{Falster2015} suggest the time is ripe for rigorous benchmarking \citep{abramowitz_towards_2012, dekauwe2014} of GVMs against empirical data.

@@

@@

Variance explained by quantitative climate variables (MAP and MAT) were tested with GAMs where variables were sequentially added to the model, and the explained variance ($R^2$) calculated. For the test of biome effects on biomass partitioning, we used a linear mixed-effects model (because two factors and their interactions were tested), again with species-dataset as the random effect. We calculated the $R^2$ for linear mixed-effects models as reported by for the fixed effects only \citep{nakagawa2013}.

@@

Figure 1a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary. We thank Martin de Kauwe for discussions and commenting on an earlier version of this paper.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
    \caption{\textbf{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning}. a) LMF for each PFT within three different biomes.  DA: deciduous angiosperms, EA: evergreen angiosperms, EG: evergreen gymnosperms. No leaf area estimates were available for boreal deciduous angiosperms. Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect. Letters denote significant differences (at $\alpha = 0.05$). b-c) Relationships between climate and LMF within each PFT. Solid lines are generalized additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
    \label{fig:figure3}
\end{figure}

<<echo=FALSE>>=
agg <- summaryBy(lalf_astba2 + lmlf_astba2 + llma ~ Group,
                   data=dataset, FUN=mean, na.rm=TRUE, keep.names=TRUE,
                   id=~pft)
agg <- subset(agg, !is.na(llma))
lm1 <- lm(lmlf_astba2 ~ llma, data=agg)
lm2 <- lm(lmlf_astba2 ~ lalf_astba2 * pft, data=agg)
r2panela <- summary(lm1)$adj.r.squared
r2panelb <- summary(lm2)$adj.r.squared
@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure4.pdf}
     \caption{\textbf{Leaf and stem traits capture variation in biomass partitioning across species.} Individual plant data were averaged by species and study combinations. Lines are linear regressions with 95\% confidence intervals. Both regressions were significant ($P \textless 0.01$). $R^2$ for fitted relationships are \Sexpr{round(100*r2panela,1)}\% in panel (a) and \Sexpr{round(100*r2panelb,1)}\% in panel (b) (for a linear regression model including PFT as a factor variable).}
  \label{fig:figure4}
\end{figure}

@@

@@

\renewcommand\thesubsection{S\arabic{subsection}}
\setcounter{subsection}{0}

\subsection{Modelling of biomass partitioning in global vegetation models (GVMs)}
\label{sec:gvmsummary}

Most leading GVMs assume that biomass partitioning between leaves and woody stems differs between major plant functional types. However, there is no consensus on how these differences are parameterized, or whether they are linked to inherent differences in leaf mass per area (LMA) or leaf lifespan (LL). We briefly summarize three classes of biomass allocation models, and how differences in biomass partitioning between PFTs arises in each of them. We base this classification on a recent model intercomparison study \cite{dekauwe2014}. See Methods \ref{sec:partterm} for use of the terms 'partitioning' and 'allocation'.

\subsubsection*{Constant allocation fractions}

Some models assume that a constant fraction of net primary production (NPP) is allocated to each component (leaves, stems, roots). CABLE and EALCO are examples of models that allocate a PFT-specific fraction of NPP to each plant component each timestep \cite{dekauwe2014} (though, for both models, alternative model formulations have been used as well). Constant allocation fractions still give rise to differences in biomass partitioning between PFTs, since turnover of leaves (determined by LL) will always differ between at least deciduous and evergreen PFTs.

\subsubsection*{Allometric constraints}

In this class of models, some allometric (or similar) relationship between biomass components is used to calculate biomass allocation in each timestep. For example, in ED2 \cite{medvigy2009}, biomass partitioning to leaves, stem sapwood and roots is constrained by a set of equations with PFT-specific parameters. Biomass allocation is then calculated at each timestep to satisfy these equations. Other models use leaf area as a constraint. For example, LPJ \cite{sitchetal2003} uses PFT-specific values of leaf area to sapwood area ratio to constrain estimates of biomass allocation. Further differences in biomass partitioning between PFTs are then the result of differences in LMA (leaf mass at a given leaf area), as well as LL (shorter lived leaves require higher annual allocation of biomass for upkeep).

\subsubsection*{Optimality}

Few GVMs use a scheme in which carbon allocation is determined by assuming biomass fractions are the result of optimization, however, this scheme is widely used in smaller-scale vegetation models \cite{valentinemakela2012,franklin2012}. The SDGVM model \cite{woodward1995} optimizes leaf area index by considering the carbon balance of the lowest leaf layer in the canopy. As with LPJ above, differences in biomass partitioning between PFTs then arise as a result of differences in LMA as well as LL.

\subsection{Biomass partitioning terminology}
\label{sec:partterm}

The terms 'partitioning' and 'allocation' have been used in various ways, confusing comparisons between studies \citep{litton2007}. Here, we define biomass partitioning as the actual distribution of biomass between compartments (e.g., leaves vs. stems), and biomass allocation as the proportion of net primary production (NPP) that is allocated to some compartment. The two processes are different because of continuous turnover of biomass, which differs strongly between compartments. We may write (cf. Ref. \citep{McMurtrieWolf1983}),

@@

\subsection{Predicting basal diameter}
\label{sec:predas}
